\documentclass[12pt]{article}
\usepackage[top=2cm,bottom=2cm,left=2cm,right=2cm]{geometry} %define borders
\usepackage{url}
\usepackage{paralist}
\usepackage{xcolor,graphicx,wrapfig,mytikz}
%\usepackage[T1]{fontenc}

\usepackage[bookmarksnumbered,bookmarksopen,colorlinks,urlcolor=gray,linkcolor=blue,citecolor=blue]{hyperref}

\usepackage{basics}

\newcommand{\defemph}[1]{\textbf{#1}}
\newcommand{\enclcite}[1]{\cite{#1}}
\newcommand{\citeByNumber}[1]{\cite{#1}}

\newcommand{\LFI}{LFI}
\newcommand{\cn}[1]{\mathtt{#1}}
\newcommand{\pn}{LATIN}

\begin{document}

\author{Florian Rabe}
\title{MMT:\\ A Foundation-Independent Approach to Formal Knowledge}
\date{\today\footnote{The latest version of this document is available at \url{https://svn.kwarc.info/repos/MMT/doc/introduction/mmt.pdf}.}}
\maketitle

\begin{abstract}
\mmt is a framework for representing declarative languages such as logics, type theories, set theories, etc..
It achieves a high level of generality by systematically avoiding a commitment to a particular syntax or semantics.
Instead, individual language features (e.g., $\lambda$-abstraction, conjunction, etc.) and syntax features (keywords, notations, etc.) are defined as separate, reusable modules, from which individual languages are assembled.
These modules can be declarative by specifying features as \mmt theories or programmatic by providing individual rules as plugins.

Despite this high degree of abstraction, it is possible to implement advanced algorithms generically at the \mmt level.
These include knowledge management algorithms (e.g, IDE, search, change management) as well as logical algorithms (e.g., parsing, type reconstruction, module system).
Thus, we can use \mmt to obtain strong implementations of declarative languages at extremely low cost.

Moreover, the focus on modularity and language-independence enables system integration, where \mmt can mediate the exchange of knowledge across different foundational systems and concrete syntaxes.
\end{abstract}

\input{main}

\section*{Acknowledgments}

The \mmt language was developed in collaboration with Michael Kohlhase.
The \mmt system includes contributions from various people at Jacobs University, most importantly Mihnea Iancu.
The results of the LATIN project were obtained in collaboration with Michael Kohlhase, Till Mossakowski (who were principal investigators), Fulya Horozal, and Mihai Codescu; the project was supported by DFG grants KO 2428/9-1 and MO 971/2-1.
The OAF project is conducted in collaboration with Michael Kohlhase; it is supported by DFG grants KO 2428/13-1 and RA-1872/3-1; the mentioned library imports were implemented in collaboration with Josef Urban and Mihnea Iancu for Mizar and with Cezary Kaliszyk for HOL Light.

\bibliographystyle{alpha}
\bibliography{../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/historical,../../../Program_Data/Latex/bib/other}

\end{document}