The main way to interact with \mmt is through the shell.
After invocation, it can be controlled via STDIN/STDOUT.
It can also be scripted by putting shell commands into files. The conventional file ending is \code{msl} for \mmt scripting language.

The \mmt shell responds with an ``empty'' \mmt instance.
Typically, the first thing to do is to load an \code{msl} that configures \mmt and registers extensions and archives.

The shell is invoked by
\begin{center}
\code{java -cp lib/*;mmt-api.jar info.kwarc.mmt.api.Run}
\end{center}
Further command line parameters are passed to the shell and executed as a command (see below). In particular,
\begin{center}
  \code{java -cp mmt-api.jar info.kwarc.mmt.api.Run file F}
\end{center}
can be used to execute a startup/configuration file \code{F}.
For both commands, Windows and Unix shell-scripts are given in the deploy folder of the \mmt repository.
In particular, users may want to asscociate the \code{msl} file ending with the \code{run-file} scripts.

The \code{-cp} parameter defines the Java classpath. Besides the main file \code{mmt-api.jar}, it must contain the dependencies, in particular the Scala library. All dependencies are available from the folder \code{deploy/lib} \footnote{Note that the classpath is a list of entries separated by \code{;} on Windows and \code{:} on Unix.}

Among the dependencies, not all files are needed all the time:
\begin{itemize}
\item If the {\mmt} web server is used, the classpath must contain \code{tiscaf.jar} \cite{tiscaf}.
\item If an SVN repository is used, the classpath must contain \code{svnkit.jar}.
\item If an {\mmt} extension is used, it (and all its dependencies) must be in the classpath.
\end{itemize}

It is often reasonable to increase the memory available to {\mmt} by using the appropriate Java parameter as in (here: 1024 MB) 
\begin{center}
\code{java -Xmx1024m -cp scala/*;mmt-api.jar info.kwarc.mmt.api.Run}
\end{center}

\subsection{General Syntax}
\emph{Commands} are given by a keyword followed by a whitespace-separated list of arguments and terminated by a newline. Empty lines and lines starting with \code{//} are ignored.

We use the following meta-variables for command arguments:
\begin{itemize}
	\item \code{F}: a file name that is interpreted relative to the current directory,
	\item \code{U}: an MMT-URI that is interpreted relative to the current base URI,
		\item \code{A}: an action that is executed on a path or object.
\end{itemize}
The shell maintains one state variable: a base MMT-URI. All paths are interpreted relative to this base.

\subsection{Basic Commands}\label{sec:shell:basic}
\begin{itemize}
\item \code{log}: Registers a log handler. The most important arguments are \code{console} to log to STDOUT and \code{file F} to log to a file.
\item \code{log+ C}, \code{log- C}: Switch on/off logging of component \code{C}.
\item \code{base U}: Sets the base path to \code{U}.
\item \code{file F}: Reads the file \code{F} and executes every line as a command. If a command causes an error, execution is aborted.
\item \code{exit}: Quits the shell.
\end{itemize}

\subsection{Interacting with Documents}\label{sec:shell:interact}
The shell stores a set of documents that are parsed into abstract data structures and made available for querying.

\begin{itemize}
\item \code{read U}: Retrieves, parses, and stores the document with URI \code{U}.
\item \code{clear}: Deletes all read knowledge items from memory.
\item \code{printAll}, \code{printXML}: Dumps the memory, used for testing.
\item Documents are accessed using actions on MMT-URIs as described below.
\end{itemize}
If the execution of these commands, requires documents that have not been read yet, these are retrieved automatically. This happens, for example, if the read document imports a theory from another document, or if the requested path points to a document that has not been read yet.

\subsection{Archive Commands}\label{sec:shell:archives}
The following commands permit the registration and manipulation of archives:

\begin{itemize}
\item \code{archive add F}: This registers an archive with local root folder \code{F}.
\item \code{archive ID O F}: This executes operation \code{O} on the archive with id \code{ID}, optionally restricted to the folder/file \code{F} (slashes as path separators). Legal values for \code{O} are \code{compile}, \code{content} (which will produce \texttt{content}, \texttt{narration}, \texttt{relational}, and \texttt{notation}), \code{mws}, \code{relational} (which reads the relational index), and \code{notation} (which reads the notation index).
\item \code{archive mar F}: This builds the \texttt{mar} file and stores it in \code{F}.
\end{itemize}

\subsection{Retrieval}\label{sec:shell:actions}
Retrieval commands provide a simple infix syntax to pipe retrieved knowledge items through some typed post-processing operations.

The command $O\;A\;a_1\;\ldots\;a_n$ evaluates $O$ and then applies a further action $A$ with additional arguments $a_i$. This corresponds to $O.A(a_1,\ldots,a_n)$ in OO-programming. These actions may be chained: $O\;A\;a_1\;\ldots\;a_n \;B\;b_1\;\ldots\;b_n$ corresponds to $O.A(a_1,\ldots,a_n).B(b_1,\ldots,b_n)$.
Below, actions are classified according to the type of $O$, which is MMT-URI, MMT-Object, or Non-MMT-Object, and the return type, which is MMT-Object, Non-MMT-object, or Nothing.
\smallskip

\noindent
Actions on MMT-URIs $U$
\begin{itemize}
\item empty action: dereferences $U$ and returns it (MMT-Object)
\item \code{closure}: dereferences $U$ and returns the closure as a self-contained document (MMT-Object)
\item \code{deps xml}: dereferences $U$ and returns its dependency set in XML representation (Non-MMT-Object)
\item \code{deps locutor}: dereferences $U$ and returns its dependency set in locutor representation (Non-MMT-Object)
\end{itemize}

\noindent
Actions on MMT-Objects $O$
\begin{itemize}
\item empty action: returns the text representation of $O$ (Non-MMT-Object)
\item \code{component C}: returns the component of $O$ called \code{C} (MMT-Object)
\item \code{xml}: returns the XML representation of $O$ (Non-MMT-Object)
\item \code{present U}: returns the rendering of $O$ using style \code{U} (Non-MMT-Object)
\end{itemize}
Valid component names \code{C} are in particular \code{type} and \code{definition} if $O$ is a constant.


\noindent
Actions on Non-MMT-Objects, all returning Nothing:
\begin{itemize}
\item empty action: prints to standard output
\item \code{write F}: prints to file \code{F}
\end{itemize}

\begin{example}
The action

\noindent
\code{U/algebra/algebra.omdoc?group closure present O/omdoc/ascii.omdoc?ascii write group.txt}
writes the presentation of the closure of the theory of groups to the file \code{group.txt} using an ASCII-based style.

\noindent
\code{U/algebra/algebra.omdoc?group?inv component type xml}
writes the type type of \code{U/algebra/algebra.omdoc?group?inv} in XML to standard output.
\end{example}

\subsection{Catalog Commands}\label{sec:shell:catalog}
The following catalog commands can be used to add catalog entries explicitly:

\begin{itemize}
%\item \code{mathpath fs F}: This is used to add local working copies to the catalog. It takes a file \code{F} in the locutor (see \url{https://locutor.kwarc.info/})  registry format and creates an entry for every working copy listed in it. The repository URLs are treated as URIs that are translated to the location of the local working copy. (See the example file \code{locutor.xml} file in the distribution.)
%\item \code{ombase F}: This creates a catalog entry for an OMBase server described in \code{F}. See the example file \code{ombase.xml} in the documentation.
\item \code{mathpath local}: This adds an entry for the local file system. URIs of the form \code{file:///U} are translated to themselves.
\item \code{mathpath fs F U}: This adds an entry for the local file system. URIs of the form \code{U/R} are translated to \code{F/R}.
\item \code{mathpath svn U}: This adds an entry for a remote SVN repository. URIs contained in it are translated to themselves.\ednote{revision}
\end{itemize}

\subsection{Server Commands}\label{sec:shell:server}
The MMT HTTP server (see Sect.~\ref{sec:http}) is controlled by the commands
\begin{itemize}
\item \code{server on P}: This starts the server on port \code{P}.
\item \code{server off}: This shuts down the server.
\end{itemize}

\subsection{GUI Commands}
The MMT GUI (see Sect.~\ref{sec:gui}) is controlled by the commands
\begin{itemize}
\item \code{browser on}: This opens a new window with an \mmt browser based Java Swing.
\item \code{browser off}: This closes the window.
\end{itemize}

\subsection{Extension Commands}\label{sec:shell:extensions}
Extensions are registered using commands of the form \texttt{kind C ARGS}. \code{C} gives the qualified class name, and \code{ARGS} is a whitespace-separated list of arguments that is passed to the extension during initialization.

\begin{itemize}
\item \code{extension C ARGS}: This registers the extension with qualified Java class name \code{C}. \code{C} must extend  \texttt{info.kwarc.mmt.api.backend.Extension}. Relevant subinterfaces of \texttt{Extension} are explained in Sect.~\ref{sec:extensions}.
All extensions take a list of string arguments, which is passed as \code{ARGS} where the arguments separated by whitespace.
\end{itemize}